---
title: "Building the SNP database"
author: "Andrew Potter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building the SNP database}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r load, message = FALSE}
library(variantCell)
```

## What this vignette needs

variantCell sits on top of two upstream tools and replaces neither:

* **cellsnp-lite**, which counts reference and alternative reads per cell at a
  list of variant positions, and
* **Vireo**, which clusters those counts into genotypes and assigns each cell to
  a genome.

Every function here consumes their output. So that this vignette actually runs,
`simulateVariantCellData()` writes a synthetic experiment in exactly the layout
a real run produces — `cellSNP.base.vcf`, `cellSNP.samples.tsv`, the AD/DP/OTH
MatrixMarket matrices, and one `donor_ids.tsv` per sample.

```{r simulate}
sim <- simulateVariantCellData(verbose = FALSE)

list.files(sim$cellsnp_paths[["S1"]])
```

Five libraries, two genetically distinct individuals in each. Two of them
(`S1`, `S2`) are repeat biopsies of the same patient, which will matter later.

```{r design}
sim$design
```

The cell metadata is an ordinary data frame whose **row names are the prefixed
cell IDs**. In a real analysis this is your Seurat or SingleCellExperiment
metadata.

```{r metadata}
head(sim$metadata, 3)
```

> The simulated data is deliberately denser than a real 10x run — roughly half
> the site-by-cell entries are non-zero, against a few percent in practice — so
> that a small example still has enough coverage to demonstrate every function.
> Cell counts and site counts are correspondingly small.

## Creating a project

```{r new-project}
project <- variantCell$new()
```

## Adding samples

`addSampleData()` reads one sample's cellsnp matrices, matches barcodes against
your cell metadata, applies the Vireo donor assignment, and stores the result.

The `donor_type` argument maps Vireo's cluster labels onto meaningful names.
**Do not hard-code it.** Vireo's `donor0` / `donor1` labels are arbitrary and
are reassigned on every run — see the donor/recipient vignette, which covers
this in full. Infer the mapping from the data instead:

```{r infer, message = FALSE, results = "hide"}
inf <- inferDonorType(sim$metadata, vireo_dir = sim$vireo_dir, verbose = FALSE)
```

```{r infer-show}
inf$donor_types[["S1"]]
inf$donor_types[["S3"]]
```

Those two are not the same mapping, which is the whole point.

```{r add-samples, results = "hide", message = FALSE}
for (s in sim$design$sample) {
  project$addSampleData(
    sample_id    = s,
    vireo_path   = sim$vireo_paths[[s]],
    cellsnp_path = sim$cellsnp_paths[[s]],
    cell_data    = sim$metadata[sim$metadata$Biopsy == s, ],
    data_type    = "dataframe",
    prefix_text  = sim$prefixes[[s]],
    donor_type   = inf$donor_types[[s]]
  )
}
```

```{r samples-added}
names(project$samples)
```

A few notes on the arguments:

* `data_type` is `"seurat"`, `"sce"` or `"dataframe"`. For a data frame,
  `vireo_path` points at the `donor_ids.tsv` **file**, not its directory, and
  row names must equal `prefix_text` + barcode.
* `prefix_text` is stripped from your cell names to recover the plain barcode
  that cellsnp-lite wrote. Derive it by removing the barcode from an existing
  cell name rather than rebuilding it from metadata columns — cell names in a
  merged object are fixed at merge time and can carry stale labels.
* For a first pass, leave `min_cells = 0` and `min_alt_frac = 0`. Filtering
  early discards sites that downstream plotting and DE need.
* `non_transplant_mode = TRUE` skips Vireo entirely, for a single-genome
  experiment.

## Building the unified database

`buildSNPDatabase()` takes the union of sites across samples, annotates them
against `EnsDb.Hsapiens.v86`, and assembles one matrix per quantity.

```{r build, results = "hide", message = FALSE}
project$buildSNPDatabase()
```

```{r build-summary}
db <- project$snp_database
dim(db$dp_matrix)
names(db)
```

Sites are **unioned**, not intersected — different samples cover different
positions, and requiring a site in every sample would discard most of them.

```{r snp-info}
head(db$snp_info, 3)
```

Annotation lands in `snp_metrics`, which is where the gene name lives:

```{r annotation}
table(db$snp_metrics$feature_type)
head(sort(table(db$snp_metrics$gene_name), decreasing = TRUE), 5)
```

The cell metadata carries everything you supplied plus the columns variantCell
adds — `sample_id`, `donor_id`, `donor_type`:

```{r cell-metadata}
colnames(db$cell_metadata)
table(db$cell_metadata$donor_type, db$cell_metadata$compartment)
```

That table is the single most important property of transplant data: the graft
supplies structural cells and the recipient supplies immune cells, almost
completely. It is what makes donor inference reliable — and it is also why
phenotypic donor-versus-recipient comparisons are largely unavailable, since the
two genomes rarely coexist within one cell type.

## Setting an identity

Most analysis functions work off a project-wide identity column:

```{r identity, results = "hide"}
project$setProjectIdentity("compartment")
```

```{r identity-show}
project$getCurrentIdentity()
```

## Adding rs IDs and population allele frequencies

If you have a reference VCF — 1000 Genomes, gnomAD — `buildSNPDatabase()` can
attach rs identifiers and population allele frequencies:

```{r rs-ids, eval = FALSE}
project$buildSNPDatabase(
  add_rs_ids        = TRUE,
  add_population_AF = TRUE,
  VCF_file_path     = "path/to/genome1K.phase3.SNP_AF5e2.chr1toX.hg38.vcf"
)
```

This chunk is not evaluated because it needs a multi-gigabyte reference file
that cannot ship with the package. Everything else in this vignette runs.

Note what such a panel implies: if you called variants against a common-variant
reference, **every site in your database is common by construction**. Nothing
rare, and nothing high-penetrance. That is a property of the panel, not of the
method — cellsnp-lite will run against any region list, which is what
`buildCellSNPDatabase()` exists to import.

## Next

* **Determining donor and recipient cells** — why the Vireo labels must be
  re-inferred every run.
* **Cell-level DE SNP analysis** — `findDESNPs()`, and what it does and does not
  measure.
* **Group-level DE SNP analysis** — `findSNPsByGroup()`, and the genotype guard
  that decides whether your contrast is answerable at all.
