---
title: "Cell-level DE SNP Analysis"
author: "Andrew Potter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cell-level DE SNP Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r load, message = FALSE}
library(variantCell)
```

## What `findDESNPs()` actually measures

Read this before using it.

`findDESNPs()` compares **read depth at a variant position** between cell
groups. Depth at a position is the abundance of the transcript containing it —
so this is an expression test, not a genotype test. Germline genotype is fixed
within an individual and cannot differ between that individual's cell types.

The measurement is real and correctly computed. The question is whether it is
the best available instrument, and mostly it is not. Measured on 52,809 cells of
a lung transplant cohort against the matched gene count matrix:

| | |
|---|---|
| reads captured at SNP positions | **51.3%** of the gene matrix |
| expressed genes with no covered SNP | **40.3%** |
| median gene: reads retained | 20% (well-covered genes up to 90%) |
| median gene: SNPs carried | **13** |
| correction burden | ~739,000 sites vs ~20,000 genes = **37× stricter** |

The coverage is not thin — half of all reads land at SNP positions. The problem
is arithmetic: per test you hold roughly 1.5% of a gene's evidence while paying
37 times the multiple-testing penalty, on 60% of the genes. **Seurat's
`FindMarkers()` on the gene count matrix strictly dominates this for "which
genes differ between these cell groups."**

Use `findDESNPs()` for:

* **A specific gene you already care about.** The multiple-testing argument
  largely evaporates once you are not scanning.
* **The coverage companion to `findSNPsByGroup()`** — confirming depth was
  adequate in the group where a variant was called absent, since otherwise
  "absent" only means "not detected".
* **Checking coverage comparability** before comparing allele fractions, since
  differential depth creates differential power.

## Setup

```{r build, results = "hide", message = FALSE}
sim <- simulateVariantCellData(verbose = FALSE)
inf <- inferDonorType(sim$metadata, vireo_dir = sim$vireo_dir, verbose = FALSE)

project <- variantCell$new()
for (s in sim$design$sample) {
  project$addSampleData(
    sample_id    = s,
    vireo_path   = sim$vireo_paths[[s]],
    cellsnp_path = sim$cellsnp_paths[[s]],
    cell_data    = sim$metadata[sim$metadata$Biopsy == s, ],
    data_type    = "dataframe",
    prefix_text  = sim$prefixes[[s]],
    donor_type   = inf$donor_types[[s]]
  )
}
project$buildSNPDatabase()
project$setProjectIdentity("compartment")
```

```{r idents}
table(project$snp_database$cell_metadata$compartment)
```

The simulator gave each gene a known depth difference between immune and
structural cells, so there is a right answer to compare against.

## Running the test

```{r de, results = "hide", message = FALSE}
res <- project$findDESNPs(
  ident.1 = "Myeloid",
  ident.2 = "Epithelial",
  n_cores = 2
)
```

`findDESNPs()` returns a **list**, not a data frame:

```{r de-structure}
names(res)
de <- res$results
dim(de)
```

```{r de-cols}
colnames(de)
```

Note the column names: `log2fc`, `pvalue`, `padj` — not the Seurat spellings.

```{r de-top}
top <- head(de[order(de$padj, -abs(de$log2fc)),
               c("gene_name", "position", "log2fc", "padj",
                 "expr_frac_1", "expr_frac_2")], 8)
top
```

## Does it recover the truth?

Aggregate to the gene level and compare against the depth shift the simulator
built in:

```{r recovery}
obs <- tapply(de$log2fc, de$gene_name, median, na.rm = TRUE)
truth <- sim$sites$immune_log2[match(names(obs), sim$sites$gene)]
cmp <- data.frame(gene = names(obs),
                  observed = round(as.numeric(obs), 2),
                  simulated = truth)
cmp <- cmp[!is.na(cmp$simulated), ]
cmp[order(cmp$simulated), ]
```

```{r recovery-cor}
round(cor(cmp$observed, cmp$simulated), 3)
```

The ordering is recovered almost exactly. Two things are worth noticing.

**The magnitudes are compressed.** With `use_normalized = TRUE` (the default),
depth is size-factor normalised and `log1p`-transformed, so a fold change on
that scale is much smaller than the underlying count ratio. Treat `log2fc` as an
ordering, not as a count ratio.

**A truly null gene does not come back at zero.** `VIM` was simulated with no
difference at all, yet reads as mildly depleted in myeloid cells. That is
compositional normalisation working correctly: immune cells carry far more
`CD74`, `FCN1` and `VCAN` reads, which inflates their size factors and pushes
everything else down. The same effect exists in ordinary RNA-seq. It is a reason
to be careful with small effects, not a bug.

**Almost everything is significant:**

```{r sig-count}
sprintf("%d of %d sites at padj < 0.05", sum(de$padj < 0.05, na.rm = TRUE), nrow(de))
```

That is what testing across ~800 individual cells buys you. Cells within a
library share a genome, a capture and a sequencing run, so they are not
independent observations; treating them as such inflates significance. This is
the standard single-cell DE pseudoreplication problem and it is the deeper
reason not to read this output as a discovery scan. When the unit of observation
matters, aggregate to the sample — `computeAlleleFractionIndex()` and
`findDEAlleleFraction()` do exactly that.

## The legitimate use: one gene you already care about

```{r targeted}
cd74 <- de[de$gene_name == "CD74", ]
sprintf("CD74: %d sites, median log2fc %.2f, %d significant",
        nrow(cd74), median(cd74$log2fc, na.rm = TRUE),
        sum(cd74$padj < 0.05, na.rm = TRUE))
```

Here you are not scanning, the gene is well covered, and the multiple-testing
objection does not apply.

## Useful arguments

```{r args, eval = FALSE}
project$findDESNPs(
  ident.1         = "Myeloid",
  ident.2         = NULL,      # NULL compares against all other cells
  donor_type      = "Donor",   # restrict to one genome
  use_normalized  = TRUE,      # size-factor normalised depth
  min_expr_cells  = 3,         # sites must be seen in this many cells
  min_alt_frac    = 0.2,       # ...at this alt fraction, to be testable
  logfc.threshold = 0.1,
  p.adjust.method = "BH",
  n_cores         = 4
)
```

`min_alt_frac` and `min_expr_cells` decide which sites are **testable**. They do
not decide which cells contribute to the effect size — that runs over all cells
in the group, following the Seurat convention, so the fold change and the
p-value describe the same population.

Alt fractions are always computed from raw AD/DP regardless of
`use_normalized`, since there is no normalised AD matrix and dividing a raw
count by a log-scaled depth does not produce a fraction.

## Visualising SNPs across cell groups

```{r heatmap, fig.width = 7, fig.height = 5}
project$plotSNPHeatmap(
  genes         = c("CD74", "SFTPC"),
  group.by      = "compartment",
  min_alt_frac  = 0,
  show_rownames = FALSE   # hundreds of sites; labels are unreadable
)
```

Immune compartments carry the `CD74` sites and structural compartments the
`SFTPC` sites, which is the depth difference the simulation built in.

To get the underlying matrix instead of the plot:

```{r heatmap-data}
hm <- project$plotSNPHeatmap(
  genes        = c("CD74"),
  group.by     = "compartment",
  min_alt_frac = 0,
  data_out     = TRUE
)
str(hm, max.level = 1)
```

## Extracting cells by SNP status

`getCellsForSNPs()` returns cell IDs meeting an allele-fraction and depth
criterion, which is how you annotate a Seurat object with variant status:

Three identifier forms are accepted: the full `CHROM_POS_REF_ALT` key, an rs ID
if you annotated against a reference VCF, and plain `chromosome:position`.

```{r cells-for-snps, results = "hide"}
info <- project$snp_database$snp_info
# pick the best-covered site in the database
best <- which.max(Matrix::rowSums(project$snp_database$dp_matrix))
target <- paste(info$CHROM[best], info$POS[best], sep = ":")

carriers <- project$getCellsForSNPs(
  snp_ids      = target,
  min_alt_frac = 0.2,
  min_dp       = 3
)
```

```{r cells-for-snps-attrs}
attr(carriers, "summary")
head(carriers[[1]], 3)
```

Those cell IDs are ready to annotate an external object:

```{r annotate-seurat, eval = FALSE}
seurat_obj$snp_status <- ifelse(colnames(seurat_obj) %in% carriers[[1]],
                                "carrier", "non-carrier")
```

The summary distinguishes cells with *any* reads at the site from cells that
passed both thresholds, which is the difference between "not carrying it" and
"not covered".

## Next

* **Group-level DE SNP analysis** — `findSNPsByGroup()` and the genotype guard.
* **Building the SNP database** — the import path.
