---
title: "Determining Donor and Recipient Cells"
author: "Andrew Potter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Determining Donor and Recipient Cells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r load, message = FALSE}
library(variantCell)
```

## The problem

Vireo separates the cells in a multiplexed library into genotype clusters and
labels them `donor0`, `donor1`, and so on. **Those labels carry no meaning.**
They are assigned independently on every run, and re-processing the same data
will happily swap them.

This is not hypothetical. Re-running a 22-sample lung transplant cohort under a
newer CellRanger version swapped the labels on **16 of 22 samples** — the same
two cell populations, opposite labels, no error and no warning.

A donor/recipient mapping recorded against one Vireo run must never be reused
against another. Doing so silently inverts Donor and Recipient, and every
downstream result inverts with it.

`inferDonorType()` re-derives the mapping from the data every time.

## How it works

The graft supplies the organ's structural tissue — epithelium, endothelium,
stroma. The recipient supplies infiltrating immune cells. So for each Vireo
cluster, compute the fraction of its cells that are structural; the higher one
is the donor.

```{r simulate}
sim <- simulateVariantCellData(verbose = FALSE)
```

The simulator randomises which genome Vireo calls `donor0`, exactly as a real
run does. Here is the ground truth it recorded:

```{r truth}
sim$truth$labels
```

Now infer it, using only the cell metadata and the Vireo output:

```{r infer}
inf <- inferDonorType(
  sim$metadata,
  vireo_dir         = sim$vireo_dir,
  sample_col        = "Biopsy",
  barcode_col       = "Barcode",
  structural_col    = "compartment",
  structural_values = c("Epithelial", "Endothelial", "Stromal")
)
```

The per-sample audit shows the separation it is working from:

```{r mapping}
inf$mapping[, c("sample", "n_donor0", "n_donor1",
                "struct_frac_donor0", "struct_frac_donor1",
                "margin", "donor0", "donor1", "call")]
```

And it recovers the truth, including the sample whose labels are flipped:

```{r validate}
truth <- sim$truth$labels
got   <- inf$mapping[match(truth$sample, inf$mapping$sample), ]
data.frame(
  sample    = truth$sample,
  simulated = ifelse(truth$donor0 == "graft", "Donor", "Recipient"),
  inferred  = got$donor0,
  agree     = (truth$donor0 == "graft") == (got$donor0 == "Donor")
)
```

The separation is normally near-binary, which is why a *low* margin is treated
as a problem rather than mild uncertainty. Samples below `min_margin` are
reported `"ambiguous"` and left `NA` rather than guessed:

```{r margin}
range(inf$mapping$margin)
```

## Using the result

`inf$donor_types` is a per-sample named vector ready to hand to
`addSampleData()`:

```{r donor-types}
inf$donor_types[["S1"]]
inf$donor_types[["S3"]]
```

```{r build, results = "hide", message = FALSE}
project <- variantCell$new()
for (s in sim$design$sample) {
  project$addSampleData(
    sample_id    = s,
    vireo_path   = sim$vireo_paths[[s]],
    cellsnp_path = sim$cellsnp_paths[[s]],
    cell_data    = sim$metadata[sim$metadata$Biopsy == s, ],
    data_type    = "dataframe",
    prefix_text  = sim$prefixes[[s]],
    donor_type   = inf$donor_types[[s]]
  )
}
project$buildSNPDatabase()
```

With the mapping correct, the composition separates as it should:

```{r composition}
db <- project$snp_database
table(db$cell_metadata$donor_type, db$cell_metadata$compartment)
```

## What getting it wrong looks like

Hard-coding `donor0 = "Donor"` for every sample produces this instead:

```{r wrong}
naive <- data.frame(
  sample   = inf$mapping$sample,
  inferred = inf$mapping$donor0,
  hardcoded = "Donor",
  result   = ifelse(inf$mapping$donor0 == "Donor", "correct", "INVERTED")
)
naive
```

```{r wrong-count}
sprintf("%d of %d samples would have donor and recipient swapped",
        sum(naive$result == "INVERTED"), nrow(naive))
```

Sixteen of twenty-two in the real cohort. The failure is silent — the analysis
runs, the plots render, and the biology is backwards for those samples.

## Confirming the clusters really are two genomes

The composition heuristic assumes Vireo split the library along genotype. That
assumption is checkable directly, without any cell-type labels:

```{r concordance}
cm <- db$cell_metadata
gc_within <- checkGenotypeConcordance(
  db,
  which(cm$sample_id == "S1" & cm$donor_type == "Donor"),
  which(cm$sample_id == "S1" & cm$donor_type == "Recipient"),
  verbose = FALSE
)
c(concordance = round(gc_within$concordance, 3), verdict = gc_within$verdict)
```

Two genomes in one library agree at only about half the callable sites. Compare
that with the same individual sampled twice — `S1` and `S2` are repeat biopsies
of the same patient:

```{r concordance-same}
gc_same <- checkGenotypeConcordance(
  db,
  which(cm$sample_id == "S1" & cm$donor_type == "Recipient"),
  which(cm$sample_id == "S2" & cm$donor_type == "Recipient"),
  verbose = FALSE
)
c(concordance = round(gc_same$concordance, 3), verdict = gc_same$verdict)
```

Near-perfect agreement, because it is one person. This is also a sample-swap
detector: the distinct-genome count comes from clustering on concordance, not
from any patient column, so a mislabelled sample cannot hide behind its
metadata.

## Other data types

`inferDonorType()` accepts a Seurat object or a SingleCellExperiment as well as
a data frame, and with `add_metadata = TRUE` returns the input object with
`vireo_donor` and `donor_type` columns attached:

```{r other-types, eval = FALSE}
inf <- inferDonorType(merged_seurat, vireo_dir = "Vireo/cellranger8")
merged_seurat <- inf$object
```

If you only want Vireo's assignments joined onto your metadata without the
donor/recipient call, the lower-level helpers do that directly:

```{r helpers, eval = FALSE}
res <- project$process_vireo_seurat(seurat_obj, vireo_path, prefix_text)
res <- project$process_vireo_sce(sce_obj, vireo_path, prefix_text)
res <- project$process_vireo_dataframe(meta_df, vireo_path, prefix_text)
```

These two chunks are not evaluated because they need your own object.

## Next

* **Building the SNP database** — the import path in full.
* **Group-level DE SNP analysis** — what a donor/recipient contrast can and
  cannot answer once you have it.
