---
title: "Group-level DE SNP Analysis"
author: "Andrew Potter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Group-level DE SNP Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r load, message = FALSE}
library(variantCell)
```

## Presence and absence is a genotype test

`findSNPsByGroup()` asks which variants are present in one group of cells and
absent in another. Because germline genotype is fixed at conception, that
question is only meaningful when the two groups are **different genomes**.

There are three regimes, and nothing in a cell-identity label tells you which
one you are in:

| your two groups are | status |
|---|---|
| two genomes, one on each side, in one library | **valid**, and strong |
| the same genome | **structurally empty** — not underpowered, empty |
| several genomes on each side | a **genetic association** design, capped by group size |

`checkGenotypeConcordance()` measures which regime applies, directly from the
data. `findSNPsByGroup()` runs it by default and warns.

## Setup

```{r build, results = "hide", message = FALSE}
sim <- simulateVariantCellData(verbose = FALSE)
inf <- inferDonorType(sim$metadata, vireo_dir = sim$vireo_dir, verbose = FALSE)

project <- variantCell$new()
for (s in sim$design$sample) {
  project$addSampleData(
    sample_id    = s,
    vireo_path   = sim$vireo_paths[[s]],
    cellsnp_path = sim$cellsnp_paths[[s]],
    cell_data    = sim$metadata[sim$metadata$Biopsy == s, ],
    data_type    = "dataframe",
    prefix_text  = sim$prefixes[[s]],
    donor_type   = inf$donor_types[[s]]
  )
}
project$buildSNPDatabase()
```

```{r design}
sim$design
```

Note that `S1` and `S2` are repeat biopsies of the same patient. That gives us
one of each regime to demonstrate.

## Measuring the regime before testing

```{r regimes}
db <- project$snp_database
cm <- db$cell_metadata

regime <- function(label, i1, i2, ...) {
  r <- checkGenotypeConcordance(db, i1, i2, verbose = FALSE, ...)
  data.frame(contrast = label,
             concordance = round(r$concordance, 3),
             genomes = paste(r$n_genomes1, "vs", r$n_genomes2),
             verdict = r$verdict)
}

rbind(
  regime("S1 Donor vs S1 Recipient",
         which(cm$sample_id == "S1" & cm$donor_type == "Donor"),
         which(cm$sample_id == "S1" & cm$donor_type == "Recipient")),
  regime("S1 vs S2 Recipient (same patient)",
         which(cm$sample_id == "S1" & cm$donor_type == "Recipient"),
         which(cm$sample_id == "S2" & cm$donor_type == "Recipient")),
  regime("all Donor vs all Recipient",
         which(cm$donor_type == "Donor"),
         which(cm$donor_type == "Recipient"),
         split_by = cm$sample_id),
  regime("Rejection vs Stable patients",
         which(cm$condition == "Rejection"),
         which(cm$condition == "Stable"),
         split_by = cm$patient)
)
```

Two genomes in one library agree at about half of callable sites. The same
person sampled twice agrees at essentially all of them. Both pooled contrasts
are flagged `heterogeneous_groups`, because each side contains several unrelated
individuals.

## Aggregating

`findSNPsByGroup()` works on aggregated counts, not on cells. There are two
modes.

**Sample-stratified** (the default) keeps samples separate within each group, so
a variant has to be consistent across samples rather than driven by one:

```{r agg-strat, results = "hide"}
agg <- project$aggregateByGroup(group_by = "condition")
```

```{r agg-strat-show}
agg$mode
agg$metadata
```

**Group-only** pools every cell in the group. Set `sample_column = NULL`:

```{r agg-group, results = "hide"}
agg_pooled <- project$aggregateByGroup(group_by = "condition",
                                       sample_column = NULL)
```

```{r agg-group-show}
agg_pooled$mode
```

Use sample-stratified for anything crossing samples. Group-only pools across
individuals, and in the real cohort its within-group concordance (0.52) was
*lower* than the between-group value (0.723) — the groups were less internally
consistent than they were different from each other.

Other useful arguments:

```{r agg-args, eval = FALSE}
project$aggregateByGroup(
  group_by            = "condition",
  sample_column       = "sample_id",
  donor_type          = "Donor",   # restrict to one genome
  min_cells_per_group = 3,
  use_normalized      = TRUE
)
```

## The valid contrast: two genomes in one library

Subset to a single sample, then split on genotype:

```{r valid, results = "hide", message = FALSE, warning = FALSE}
s1 <- project$subsetVariantCell(column = "sample_id", values = "S1")
agg_s1 <- s1$aggregateByGroup(group_by = "donor_type")
res_s1 <- s1$findSNPsByGroup(
  ident.1 = "Donor",
  ident.2 = "Recipient",
  aggregated_data = agg_s1
)
```

```{r valid-verdict}
res_s1$genotype_check$verdict
c(sites_tested = res_s1$summary$total_tested,
  sites_passing = res_s1$summary$passed_filters)
table(res_s1$results$presence)
```

Hundreds of variants are present in one genome and absent in the other. This is
the regime the function is for, and the answer is a genotype call rather than a
statistical claim — with one library per side there is no sample-level
replication to compute a p-value from. Rank on `presence_score`:

```{r valid-top}
top <- head(res_s1$results[order(-res_s1$results$presence_score),
                           c("chromosome", "position", "gene_name",
                             "alt_frac_1", "alt_frac_2", "presence",
                             "presence_score")], 6)
top
```

## The capped contrast: patients against patients

```{r capped, results = "hide", message = FALSE}
res_cond <- withCallingHandlers(
  project$findSNPsByGroup(ident.1 = "Rejection", ident.2 = "Stable",
                          aggregated_data = agg),
  warning = function(w) invokeRestart("muffleWarning")
)
```

The guard travels with the result, so a flagged contrast is still recognisable
when the object is read back months later:

```{r capped-check}
g <- res_cond$genotype_check
data.frame(
  verdict              = g$verdict,
  genomes              = paste(g$n_genomes1, "vs", g$n_genomes2),
  min_attainable_p     = signif(g$min_attainable_p, 3),
  bonferroni_threshold = signif(g$bonferroni_threshold, 3),
  reachable            = g$significance_reachable
)
```

```{r capped-results}
res_cond$summary$significant_results
```

Zero, and no amount of extra sequencing changes that.

Note that the guard reports **2 vs 2**, not the three-against-two you might
expect from the sample counts. The independent unit is the individual, not the
library — `S1` and `S2` are the same patient, so they contribute one genome
between them. The count comes from clustering on genotype concordance rather
than from any patient column, so it cannot be inflated by a mislabelled sample.

With two individuals against two, the smallest two-sided p a **perfectly
separating** variant can produce is `2 / choose(4, 2)` = 0.333, against a
Bonferroni threshold of roughly 2.5e-5. The ceiling depends only on the group
sizes:

```{r ceiling}
ceiling_p <- function(n1, n2) 2 / choose(n1 + n2, n1)
data.frame(
  individuals = c("2 vs 2", "5 vs 4", "10 vs 10", "14 vs 14"),
  best_attainable_p = signif(c(ceiling_p(2, 2), ceiling_p(5, 4),
                               ceiling_p(10, 10), ceiling_p(14, 14)), 3)
)
```

Against ~2,000 sites here — or ~739,000 in a real database — you need on the
order of fourteen versus fourteen individuals before a perfect separator clears
correction at all, and real association studies run to hundreds because nothing
separates perfectly.

This design is not *invalid*; it is a case-control genetic association study,
which is a real thing. It simply cannot reach significance at this n, so **rank
hits as exploratory candidates rather than thresholding them**. Two caveats
survive at any n and ranking does not fix either: unrelated individuals differ
in ancestry, which shifts allele frequencies genome-wide independently of your
grouping variable; and if your groups differ in 10x chemistry, the two arms
interrogate different variant space.

## The empty contrast

```{r empty, results = "hide", message = FALSE}
p1 <- project$subsetVariantCell(column = "patient", values = "P1")
agg_p1 <- p1$aggregateByGroup(group_by = "sample_id")
```

```{r empty-check}
cm1 <- p1$snp_database$cell_metadata
gc <- checkGenotypeConcordance(
  p1$snp_database,
  which(cm1$sample_id == "S1"),
  which(cm1$sample_id == "S2"),
  verbose = FALSE
)
c(concordance = round(gc$concordance, 3), verdict = gc$verdict)
```

`same_genome`. One person's germline genotype does not change between biopsies,
so there is nothing here to find and any apparent hit is transcript abundance or
coverage. If that is what you want to test, use `findDESNPs()` — or better,
ordinary gene-level differential expression.

## Reading the results

```{r results-cols}
colnames(res_s1$results)
```

The columns worth knowing:

* `alt_frac_1` / `alt_frac_2` — alternative allele fraction in each group.
* `presence` — which group the variant was called present in.
* `presence_score` — composite of alt-fraction difference, sample consistency
  and coverage. This is the ranking statistic.
* `sample_consistency` — fraction of samples in the group agreeing. Meaningless
  when a group has one sample.
* `p_value` / `p_adjusted` / `significant` — only interpretable when the design
  can actually reach significance. Check `genotype_check` first.
* `depth_1` / `depth_2` — summed depth. Note this is `rowSums` **across
  samples**, so `min_depth = 10` over eleven samples is a weak per-sample
  guarantee.

```{r save, eval = FALSE}
write.csv(res_s1$results, "SNPs_donor_vs_recipient_S1.csv", row.names = FALSE)
```

## Visualising SNPs in a gene

```{r plot-basic, fig.width = 7, fig.height = 4.5}
project$plotSNPs(
  "CD74",
  group.by     = "compartment",
  min_alt_frac = 0,
  plot_density = FALSE
)
```

Point position is genomic coordinate, colour is alternative allele fraction, and
size is read depth.

Split within each group to compare genotypes side by side:

```{r plot-split, fig.width = 7, fig.height = 5.5}
project$plotSNPs(
  "CD74",
  group.by     = "condition",
  split.by     = "donor_type",
  min_alt_frac = 0.2,
  plot_density = TRUE
)
```

Or take the underlying data frame instead of the plot:

```{r plot-data}
jak1 <- project$plotSNPs(
  "JAK1",
  group.by     = "compartment",
  min_alt_frac = 0,
  data_out     = TRUE
)
head(jak1, 4)
```

`plotSNPs()` looks the gene up in `EnsDb.Hsapiens.v86`, so it needs a symbol
that resolves to exactly one record. Genes with alternative-haplotype or
readthrough entries — `HLA-A`, `B2M`, `PTPRC` — return several and will not
plot.

## Testing allele fraction instead

If your question is about the *fraction* of reads carrying the alternative base
rather than presence or absence — RNA editing rate, allele-specific expression,
mitochondrial heteroplasmy — that is a different statistic, and
`R/09-allele-fraction.R` provides it. The unit of observation is the sample, not
the cell:

```{r allele-fraction}
idx <- project$computeAlleleFractionIndex(group_by = "donor_type")
head(idx, 4)
```

One index per pseudobulk means one statistic, so there is no genome-wide
correction to pay — which is why it works at cohort sizes where a per-site scan
cannot. `findDEAlleleFraction()` tests sites individually and reports the same
attainable-p ceiling before you commit to it.

## Next

* **Cell-level DE SNP analysis** — `findDESNPs()`, and why it is not a discovery
  scan.
* **Determining donor and recipient cells** — getting the genotype labels right
  in the first place.
